---
title: "Full Mouse Gastrulation Dataset"
author: "Lindsay Hayes"
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
  html_document:
    toc: yes
    df_print: paged
editor_options:
  chunk_output_type: inline
---

## Load Packages
```{r message=FALSE}
library(scran)
library(scater)
library(tidyverse)
```


## Load Full Gastrulation Dataset
This will take a few minutes to load so be patient. 
You need to run all lines to log normalize the data.
It will take several minutes.
```{r}
load("~/workspace/c_moor_data/gastrulation/gastrulation_sce.rda")
sce <- sce[,colData(sce)$stage != "mixed_gastrulation"]
sce <- logNormCounts(sce)
```

In your previous gatrulation dataset you only had 1 single sample that was a snapshot of the full dataset. Last week you went over the different types of metadata in the gastrulation dataset including: `celltype`, `stage`, and `sample`

Now with the full data set you can perform your `plotReducedDim` and `plotExpression` using `stage`! Try it out.

# Plot cells from all the samples.
For your project, use this sce data along with your prevous R Notebook to plot by stage or sample.
```{r}
plotReducedDim(object = sce, dimred="umap", colour_by="stage")
```
# Plot the expression of your gene of interest
```{r}
gene_name <- "Wnt1"
gene_id <- rowData(sce)[which(rowData(sce)[,2] %in% gene_name),1]
plotExpression(object = sce, features = gene_id, x = "stage", colour_by = "stage") + theme(axis.text.x = element_text(angle = 90)) + labs(title = gene_name)
```

Try one (or more!) of the following tricks for your project work this week. Feel free to post on discourse if you have problems or want feedback on a plot!

## Check out: 

https://help.c-moor.org/c/look-at-this/8

https://help.c-moor.org/t/tips-for-pretty-plots/192
