---
title: "Explore Mouse Gastrulation Metadata"
author: "Lindsay Hayes"
date: "`r Sys.Date()`"
output:
  html_document: 
    toc: yes
    df_print: tibble
  html_notebook: 
    toc: yes
    toc_float: yes
editor_options:
  chunk_output_type: inline
---

# Summary

- Embryo Development -- https://www.pnas.org/content/105/34/12331
- Publication -- https://pubmed.gov/30787436
- Shiny -- https://marionilab.cruk.cam.ac.uk/MouseGastrulation2018
- Dataset -- https://bioconductor.org/packages/MouseGastrulationData

# Load packages

```{r message=FALSE}
library( "tidyverse" )
```

# Check out Cell Metadata
```{r, show_col_types = FALSE}
df_meta <- read_tsv( "/home/idies/workspace/c_moor_data/gastrulation/dev/gastrulation_meta.tsv" )
df_meta <- df_meta[df_meta$stage != "mixed_gastrulation",]
df_meta <- df_meta[!is.na(df_meta$celltype),] 
colnames(df_meta)
head(df_meta)
```

# Lets explore the metadata:  sample, stage, celltype
## Number of cells of each celltype
```{r}
table( df_meta$celltype )
par(mar=c(5,10,1,1))
barplot(table( df_meta$celltype ), las = 1, horiz=T, cex.names=0.5, main = "Cells per Celltype", xlab="number of cells")
```

## Number of cells at each embryonic developmental stage
```{r}
table( df_meta$stage )
barplot(table( df_meta$stage ), las = 1, horiz=T, cex.names=0.5, main = "Cells per Stage", xlab="number of cells")
```

## Number of cells from each sample
```{r}
table( df_meta$sample )
barplot(table( df_meta$sample ), las = 1, horiz=T, cex.names=0.5, main = "Cells per Sample", xlab="number of cells")
```

## How many cells come from each sample at each stage?
```{r}
table(df_meta$sample, df_meta$stage,dnn = c("sample", "                stages"))
# big messy table with lots of zeros we don't need!
```
```{r}
# We can clean up the table with group_by and summarise
df_meta %>% group_by(stage) %>% summarise(samples = length(unique(sample)), cells = length(sample))
```
```{r}
barplot(table(df_meta$sample, df_meta$stage), main = "cells per sample per stage", ylab = "number of cells", col=rainbow(37), legend=TRUE,args.legend=list(cex=0.5,x="topleft", ncol = 3))
```

### Why might there be a differences in the number of cells from each sample/stage? 

## How many cells come from each stage from each celltype?
```{r}
df_meta %>% group_by(celltype) %>% summarise(stages = length(unique(stage)), cells = length(stage))
```

```{r}
cbind(c(1:38), table(df_meta$celltype, df_meta$stage))[,1:2]
```

```{r}
par(mar=c(4, 7, 1, 1))
barplot(table(df_meta$stage, df_meta$celltype), las = 1, horiz=T, cex.names=0.5, main = "Stages per cell type", xlab="number of cells", col=rainbow(10), legend=TRUE,args.legend=list(cex=0.7,x="topright", ncol = 2))
```

## What are the trends for cell types changing across stage? 
```{r}
# how does a celltype change over time?
barplot(table( df_meta$celltype, df_meta$stage )[32,], main = "Primitive Streak", ylab = "number of cells")
barplot(table( df_meta$celltype, df_meta$stage )[11,], main = "Epiblast", ylab = "number of cells")
```

### What about the notochord or mesenchyme?

## What is the composition of cell types at each stage?
```{r}
cbind(c(1:9), table(df_meta$stage, df_meta$celltype))[1:9, 1:4]

barplot(table( df_meta$stage, df_meta$celltype )[3,], main = "Cell type diversity at E7.0", ylab = "number of cells", las=2, cex.names = 0.7)
```

### Try E8.5 or E6.5

### Look at **abundance** of each cell type across stages
```{r}
table(df_meta$stage)
table(df_meta$stage, df_meta$celltype)[1:9, 1:4]
round(as.matrix(table(df_meta$stage, df_meta$celltype))/as.vector(table(df_meta$stage))*100,1)[1:9,1:4]
abundance <- t(round(as.matrix(table(df_meta$stage, df_meta$celltype))/as.vector(table(df_meta$stage))*100,1))
```
```{r}
par(mar=c(5,7,1,1))
barplot(abundance[,"E6.5"], las = 1, horiz=T, cex.names=0.5, main= "E6.5", xlab = "percent of cells")
```

### What about a later stage?

# Document software
```{r}
sessionInfo()
```